---
title: "Graphs and Maps for MTA Income"
author: "Henry Kanengiser"
date: "2023-01-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(sf)
library(ggplot2)

# DIRECTORY
wd <- getwd()
csv <- paste0(wd, "/csv")
dss <- paste0(wd, "/dat/station_summary")
dln <- paste0(wd, "/Subway_Lines")
dbo <- paste0(wd, "/dat/nyboroughs_22b")

# FILE
fil_sum <- "trainlineincome.csv"
fil_statsum <- "station_summary.shp"
fil_lines <- "geo_export_6a96c007-c8ad-4c2f-9447-d3927130ac22.shp"
fil_boros <- "nybb.shp"

```

The **purpose** of this program is to produce graphs and static maps based on the subway line analysis. This program also sets up a prepared dataset that can be used in the R Shiny app to produce an interactive map.

# Read in summary files
```{r}

## Station summmary shapefile
stations <- st_read(dsn = file.path(dss, fil_statsum))

## Line summary dataframe
linesum <- read_csv(file.path(csv, fil_sum))

## Line shapefiles
lineshp <- st_read(dsn = file.path(dln, fil_lines))

## Borough shapefile
boroughs <- st_read(dsn = file.path(dbo, fil_boros)) %>%
  st_transform(2263)

```

Create color palette for subway lines
```{r}

rs_group_pal <- c(
  "#0A2F51", # None OR No to 25% Rent Stabilized when used with rent_stab_ariel
  "#1D9A6C", # Some OR 26-75% Rent Stabilized when used with rent_stab_ariel 
  "#81CC3D"  # All OR 75% to Fully Rent Stabilized when used with rent_stab_ariel
)

subway_pal <- c(
  "#ee352e", #1/2/3
  "#00933c", #4/5/6
  "#b933ad", #7
  "#0039a6", #A/C/E
  "#ff6319", #B/D/F/M
  "#6cbe45", #G
  "#996633", #J/Z
  "#a7a9ac", #L
  "#808183", #S
  "#fccc0a"  #N/Q/R/W
)

```


## Plot income by train line

```{r}
linesum %>%
  ggplot() + 
  aes(x = forcats::fct_reorder(line, mhhi21, min, .desc = TRUE), y = mhhi21) + 
  geom_col() + 
  labs(
    title = "Median household income along each subway line",
    caption = "Note: Income is shown in 2021 dollars",
    x = NULL,
    y = NULL
  ) + 
  scale_y_continuous(labels = scales::dollar) + 
  theme_minimal()
  
```

### Compare line income over time

```{r}

line_income_long <- linesum %>%
  select(line, mhhi16, mhhi21) %>%
  # inflation adjust from 2016 to 2021 dollars
  mutate(mhhi16 = mhhi16 * 1.2417,
         changehhi = mhhi21-mhhi16) %>%
  mutate(line = forcats::fct_reorder(line, mhhi21, min, .desc = TRUE)) %>%
  pivot_longer(cols = -c(line, changehhi), names_to = "year", names_prefix = "mhhi", values_to = "inc") %>%
  mutate(year = ifelse(year == "16", "2012-2016", "2017-2021"))


line_income_long %>%
  ggplot() + 
  aes(x = line, y = inc, fill = year) + 
  geom_col(position = "dodge") + 
  labs(
    title = "Median household income along each subway line over the past 10 years",
    caption = "Note: Income is shown in 2021 dollars",
    x = NULL
  ) + 
  scale_y_continuous(labels = scales::dollar) + 
  theme_minimal()

```


# Maps

First, prep the line shapefile to create maps
```{r}

lineshp %>% 
  st_drop_geometry() %>%
  count(rt_symbol, name)

# Add colors to line files
lineshp2 <- lineshp %>%
  mutate(color = case_when(
    rt_symbol == "1" ~ "#ee352e",
    rt_symbol == "4" ~ "#00933c",
    rt_symbol == "7" ~ "#b933ad",
    rt_symbol == "A" ~ "#0039a6",
    rt_symbol == "B" ~ "#ff6319",
    rt_symbol == "G" ~ "#6cbe45",
    rt_symbol == "J" ~ "#996633",
    rt_symbol == "L" & name == "L" ~ "#a7a9ac",
    rt_symbol == "L" & name == "S" ~ "#808183",
    rt_symbol == "N" ~ "#fccc0a") %>%
      factor(levels = c("#ee352e", "#00933c", "#b933ad",
                        "#0039a6", "#ff6319", "#6cbe45",
                        "#996633", "#a7a9ac", "#808183",
                        "#fccc0a" ))
  )

# lineshp2 %>%
#   st_drop_geometry() %>%
#   count(color, rt_symbol)

```

```{r}

ggplot() + 
  geom_sf(data = boroughs, mapping = aes(fill = NULL)) +
  geom_sf(data = lineshp2, mapping = aes(color = color), show.legend = FALSE) + 
  scale_color_manual(values = subway_pal) + 
  geom_point(data = stations, mapping = aes(size = mhhi21, geometry = geometry), stat = "sf_coordinates") + 
  coord_sf(crs = st_crs(2263))

```

